---
title: "NMDS and PERMANOVA"
subtitle: "Supporting code for manuscript:Comprehensive genome analysis of Neisseria meningitidis from South America reveals a distinctive pathogenicity-related prophages repertoire"
author: "Benjamin Leyton-Carcaman | BIOREN-UFRO | <leyton.bl@gmail.com>"
date: "30-06-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Non-parametric multidimensional scaling (NMDS) analysis.

NMDS is a non-parametric version of PcoA. NMDS represents better than PcoA the variation between groups. More information [here](https://riffomonas.org/code_club/2022-02-14-nmds)

#### import libraries

```{r message=FALSE, warning=FALSE}
packages<-c("tidyverse", "vegan", "readxl", "ggplot2", "writexl", "ggsci", "patchwork")

.inst <- packages %in% installed.packages()
if(any(!.inst)) {
  install.packages(packages[!.inst])
}

sapply(c(packages), require, character.only = TRUE)
```

### import datasets

#### Genogroup

MDA is included for reference, however its effects are not significant in this study

```{r warning=FALSE}

datos<-read_excel("Matrices.xlsx", sheet = 6)

datos_genogroup<-datos[,-c(1,2,4,5,6)]
datos_genogroup<-na.omit(datos_genogroup)

datos_genogroup %>%
  group_by(genogroup) %>%
  mutate(n=n()) %>%
  filter(n>4) -> datos_genogroup

datos_genogroup<-datos_genogroup[,-59]

geno_dist<-vegdist(datos_genogroup[,-1], method='jaccard')
set.seed(2021)

#permanova
geno_div<-adonis2(geno_dist~genogroup, data=datos_genogroup, permutations = 999, method="jaccard")
set.seed(30)

#ordinacion

genoMDS<-metaMDS(datos_genogroup [,-1], k=2, trymax=1000, autotransform=TRUE)

NMDS1_geno<- genoMDS$points[,1]
NMDS2_geno<- genoMDS$points[,2]
geno.plot<-cbind(datos_genogroup, NMDS1_geno=NMDS1_geno, NMDS2_geno=NMDS2_geno)

#vectores

fit_geno<-envfit(genoMDS, datos_genogroup)
arrow_geno<-data.frame(fit_geno$vectors$arrows,R = fit_geno$vectors$r, P = fit_geno$vectors$pvals)


arrow_geno$FG <- rownames(arrow_geno)
arrow.p_geno<-filter(arrow_geno, P <= 0.05)
arrow.p_geno<-filter(arrow_geno, R >= 0.3)

pp_geno<-ggplot(data=geno.plot, aes(NMDS1_geno, NMDS2_geno))+
  geom_point(data=geno.plot, aes(NMDS1_geno, NMDS2_geno, color=genogroup),position=position_jitter(.2), size=2)+
  stat_ellipse(aes(fill=genogroup), alpha=.2,type='t', geom="polygon")+ 
  stat_ellipse(aes(color=genogroup),type = "t", linetype = 1, alpha=1, size=1.0) +
  theme_classic ()+
  geom_segment(data=arrow.p_geno, aes(x=0, y=0, xend=NMDS1*R*3, yend=NMDS2*R*3), arrow=arrow(length=unit(0.2, "cm"))) +
  geom_label(data = arrow.p_geno, aes(x=NMDS1*R*3, y=NMDS2*R*3, label=FG), position = "identity")

pp_geno
geno_div

```

#### Clonal-Complex

MDA is included for reference, however its effects are not significant in this study

```{r warning=FALSE}

com<-read_excel("Matrices.xlsx", sheet = 6)
complejo_clonal<-com[,-c(1:3,5,6)]
complejo_clonal<-na.omit(complejo_clonal)

complejo_clonal %>%
  group_by(`Clonal-complex`) %>%
  mutate(n=n()) %>%
  filter(n>4) -> complejo2

complejo_st<-as.data.frame(complejo2[,-59])


st_dist<-vegdist(complejo_st[-1], method = "jaccard")

#PERMANOVA

st_div<-adonis2(st_dist~`Clonal-complex`, complejo_st, permutations = 999, method = "jaccard")
set.seed(2022)

st_nmds<-metaMDS(complejo_st[-1], k=2, trymax=1000, autotransform=TRUE)

NMDS1_st<-st_nmds$points[,1]
NMDS2_st<-st_nmds$points[,2]

st.plot<-cbind(complejo_st,NMDS1=NMDS1_st, NMDS2=NMDS2_st)

#vectores


fit_st<-envfit(st_nmds, complejo_st)
arrow_st<-data.frame(fit_st$vectors$arrows,R = fit_st$vectors$r, P = fit_st$vectors$pvals)
arrow.p_st<-filter(arrow_st, P <= 0.05)
arrow.p_st<-filter(arrow_st, R >= 0.4)



arrow_st$FG <- rownames(arrow_st)

arrow.mda<-arrow_st[12,]
arrow.p_st<-bind_rows(arrow.mda, arrow.p_st)

#grafico

colores_st<-c("#38761d","#e3359b","#ea9999", "#f1c232","#45818e", "#274e13", "#b4a7d6", "#f6b26b", "#a2c4c9")
pp_st<-ggplot(data=st.plot, aes(NMDS1, NMDS2))+
  geom_point(data=st.plot, aes(NMDS1, NMDS2, color=`Clonal-complex`),position=position_jitter(.2), size=2)+
  stat_ellipse(aes(fill=`Clonal-complex`), alpha=.2,type='t',size =3, geom="polygon")+
  stat_ellipse(aes(color=`Clonal-complex`),type = "t", linetype = 1, alpha=1, size=1.0) +
  theme_classic()+ geom_segment(data=arrow.p_st, aes(x=0, y=0, xend=NMDS1*R*3, yend=NMDS2*R*3), arrow=arrow(length=unit(0.2, "cm"))) + 
  geom_label(data = arrow.p_st, aes(x=NMDS1*R*3, y=NMDS2*R*3, label=FG, size=10), position = "identity") +
  scale_color_manual(values = colores_st) + scale_fill_manual(values=colores_st) + theme(legend.position = "right", legend.title = element_blank(), legend.text = element_text(size = 15),legend.key = element_rect(fill = "white", colour = "black"))


pp_st
st_div
arrow.p_st

```

#### Country

MDA is included for reference, however its effects are not significant in this study

```{r}

datos_country<-datos[,-c(1:4,6)]
datos_country<-na.omit(datos_country)


country_dist<-vegdist(datos_country[,-1], method='jaccard')
set.seed(2021)

#permanova
country_div<-adonis2(country_dist~Country, data=datos_country, permutations = 999, method="jaccard")
set.seed(30)

#ordinacion

countryMDS<-metaMDS(datos_country [,-1], k=2, trymax=1000, autotransform=TRUE)

NMDS1_country<- countryMDS$points[,1]
NMDS2_country<- countryMDS$points[,2]
country.plot<-cbind(datos_country, NMDS1_country, NMDS2_country)

#vectores

fit_country<-envfit(countryMDS, datos_country)
arrow_country<-data.frame(fit_country$vectors$arrows,R = fit_country$vectors$r, P = fit_country$vectors$pvals)


arrow_country$FG <- rownames(arrow_country)
arrow.p_country<-filter(arrow_country, P <= 0.05)
arrow.p_country<-filter(arrow_country, R >= 0.3)

#grafico

pp_country<-ggplot(data=country.plot, aes(NMDS1_country, NMDS2_country))+
  geom_point(data=country.plot, aes(NMDS1_country, NMDS2_country, color=Country),position=position_jitter(.2), size=2)+
  stat_ellipse(aes(fill=Country), alpha=.2,type='t',size =3, geom="polygon")+ 
  stat_ellipse(aes(color=Country),type = "t", linetype = 1, alpha=1, size=1.0) +
  theme_classic()+
  geom_segment(data=arrow.p_country, aes(x=0, y=0, xend=NMDS1*R*3, yend=NMDS2*R*3), arrow=arrow(length=unit(0.2, "cm"))) +
  geom_label(data = arrow.p_country, aes(x=NMDS1*R*3, y=NMDS2*R*3, label=FG), position = "identity")

pp_country
country_div
arrow.p_country


```

#### pathogenesis

MDA is included for reference, however its effects are not significant in this study

```{r message=FALSE, warning=FALSE}

matrix <- read_excel("./Matrices.xlsx", sheet = "matrix04")
df<-matrix[,c(2,3:58)]

disease <- matrix[,1:2]
nm.matrix<-as.matrix(df[-1])
disease$pathogenesis<-as.factor(disease$pathogenesis)

nm.dist<-vegdist(nm.matrix, method='jaccard')
set.seed(2021)
nm.div<-adonis2(nm.dist~pathogenesis, data=matrix, permutations = 999, method="jaccard")
nm.div

nmMDS<-metaMDS(nm.matrix, distance="jaccard", k=2, trymax=1000, autotransform=TRUE)
nmMDS

NMDS1<- nmMDS$points[,1]
NMDS2<- nmMDS$points[,2]
nm.plot<-cbind(nm.matrix, NMDS1, NMDS2, disease)

fit<-envfit(nmMDS, nm.matrix)
arrow<-data.frame(fit$vectors$arrows,R = fit$vectors$r, P = fit$vectors$pvals)


arrow$FG <- rownames(arrow)
arrow.p<-filter(arrow, P <= 0.05)
arrow.R<-filter(arrow, R >= 0.4)
mda<-rbind(arrow.R, arrow.p[9,])

#grafico

pqq<-ggplot(data=nm.plot, aes(NMDS1, NMDS2))+
  geom_point(data=nm.plot, aes(NMDS1, NMDS2, color=pathogenesis, shape=pathogenesis),position=position_jitter(.2), size=2)+
  stat_ellipse(aes(fill=pathogenesis), alpha=.2,type='t',size =1, geom="polygon")+ 
  stat_ellipse(aes(color=pathogenesis),type = "t", linetype = 1, alpha=1, size=1.0) +
  theme_classic()+
  geom_segment(data=mda, aes(x=0, y=0, xend=NMDS1*R*4, yend=NMDS2*R*4), arrow=arrow(length=unit(0.2, "cm"))) +
  geom_label(data = mda, aes(x=NMDS1*R*4, y=NMDS2*R*4, label=FG, size=10), position = "identity") +
  scale_color_hue(direction = -1) + scale_fill_hue(direction = -1) + theme(legend.position = "right", legend.title = element_blank(), legend.text = element_text(size = 15),
                                                                           legend.key = element_rect(fill = "white", colour = "black"))

pqq
```

